Early Warning Signals and the Prosecutor's Fallacy

### 2012-06-20 fallacy.tex
formatting for PRSB submission

@@

\begin{figure}
  \begin{center}
    \subfigure[\textbf{True criminal} A bifurcation
    drives a transition into the undesirable state]{
    \includegraphics[width=.4\textwidth]{bifurcation.pdf}
    \label{fig:bifurcation} } \subfigure[\textbf{Innocent suspect} A
    transition to the alternative stable state has occurred purely by
    chance.]{ \includegraphics[width=.4\textwidth]{fallacytransition.pdf}
    \label{fig:fallacy} }
  \end{center} \caption{\textbf{The Prosecutor's Fallacy}. Plots of system
  state versus potential in two cases.  Potential is the negative integral
  of the mean of dynamical equation.  The potential function gives an
  intuitive picture of the stability of a system by imagining the curve as
  a surface on which a ball is free to bounce across -- wells correspond
  to stable points and peaks to unstable points.  Early warning signals
  are aimed at detecting systems which are slowly moving towards a tipping
  point or bifurcation, illustrated in the successively darker potential
  curves in panel~\subref{fig:bifurcation}.  An example trajectory from
  a simulation under this process shows the state of the system as the
  potential moves towards the bifurcation point. The state of the system
  in the simulation is given by the horizontal axis, while the vertical
  dimension is used to show time on the right-hand axis.  Some stable
  systems, panel~\subref{fig:fallacy}, may experience chance transitions
  to an alternate stable state without the presence of this slow change
  in the shape of the potential well which tips the system off to the left.
  An example of the trajectory of such a chance transition is
  plotted over the potential.  While such events may be unlikely over
  short time-scales, selecting historical examples of systems that
  have experienced sudden transitions increases the probability of
  accidentally using chance events depicted in panel~\subref{fig:fallacy}
  to test methods designed to detect the kind of process depicted in
  panel~\subref{fig:bifurcation}.} \label{fig:1}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=6in]{indicators.pdf}
  \end{center}
  \caption{The distribution of the correlation statistic $\tau$ for two
  early warning indicators (variance, autocorrelation) on replicates
  conditionally selected for having collapsed by chance in simulations
  is shown in grey bars.  Solid lines indicate the estimated density of
  the statistic from a random sample of the simulations (not conditional
  on observing a transition). Positive values of $\tau$ correspond to
  a pattern of an indicator increasing with time; typically taken as
  evidence that a system is approaching a critical transition.  In these
  simulations, the pattern arises instead from the Prosecutor's fallacy
  of conditional selection.}
  \label{fig:indicator}
\end{figure}

@@

 \section{References}%bibliography
 \bibliography{fallacy}
 \end{document}

>>>>

\usepackage{endfloat} % Move floats to the end of the document
\renewcommand{\listoffigures}{} % but supress these lists
\renewcommand{\listoftables}{} % supress these lists
% compile without figures?
\newif\ifhavefigures
%\havefigurestrue
\newif\ifnohavefigures % allows alternate condition
\nohavefigurestrue     % turn on the alternate condition uncommenting this

@@

\begin{figure}
  \begin{center}
    \subfigure[\textbf{True criminal} A bifurcation
      drives a transition into the undesirable state]{
      \ifhavefigures
%        \includegraphics[width=.4\textwidth]{figure1a}
      \fi
      \ifnohavefigures
        \hspace{.4\textwidth}
      \fi
      \label{fig:bifurcation}
    } 
    \subfigure[\textbf{Innocent suspect} A
    transition to the alternative stable state has occurred purely by
    chance.]{
    \ifhavefigures
%      \includegraphics[width=.4\textwidth]{figure1b}
    \fi
    \ifnohavefigures
      \hspace{.4\textwidth}
    \fi
    \label{fig:fallacy} }
  \end{center} \caption{\textbf{The Prosecutor's Fallacy}. Plots of system
  state versus potential in two cases.  Potential is the negative integral
  of the mean of dynamical equation.  The potential function gives an
  intuitive picture of the stability of a system by imagining the curve as
  a surface on which a ball is free to bounce across -- wells correspond
  to stable points and peaks to unstable points.  Early warning signals
  are aimed at detecting systems which are slowly moving towards a tipping
  point or bifurcation, illustrated in the successively darker potential
  curves in panel~\subref{fig:bifurcation}.  An example trajectory from
  a simulation under this process shows the state of the system as the
  potential moves towards the bifurcation point. The state of the system
  in the simulation is given by the horizontal axis, while the vertical
  dimension is used to show time on the right-hand axis.  Some stable
  systems, panel~\subref{fig:fallacy}, may experience chance transitions
  to an alternate stable state without the presence of this slow change
  in the shape of the potential well which tips the system off to the left.
  An example of the trajectory of such a chance transition is
  plotted over the potential.  While such events may be unlikely over
  short time-scales, selecting historical examples of systems that
  have experienced sudden transitions increases the probability of
  accidentally using chance events depicted in panel~\subref{fig:fallacy}
  to test methods designed to detect the kind of process depicted in
  panel~\subref{fig:bifurcation}.} \label{fig:1}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \ifhavefigures
%    \includegraphics[width=6in]{figure2}
    \fi
  \end{center}
  \caption{The distribution of the correlation statistic $\tau$ for two
  early warning indicators (variance, autocorrelation) on replicates
  conditionally selected for having collapsed by chance in simulations
  is shown in grey bars.  Solid lines indicate the estimated density of
  the statistic from a random sample of the simulations (not conditional
  on observing a transition). Positive values of $\tau$ correspond to
  a pattern of an indicator increasing with time; typically taken as
  evidence that a system is approaching a critical transition.  In these
  simulations, the pattern arises instead from the Prosecutor's fallacy
  of conditional selection.}
  \label{fig:indicator}
\end{figure}

@@

 \section{References}%bibliography

 \begin{thebibliography}{24}
\expandafter\ifx\csname natexlab\endcsname\relax\def\natexlab#1{#1}\fi
\expandafter\ifx\csname url\endcsname\relax
  \def\url#1{\texttt{#1}}\fi
\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi

\bibitem[{Berkes et~al.(2006)Berkes, Hughes, Steneck, Wilson, Bellwood, Crona,
  Folke, Gunderson, Leslie, Norberg, Nystr\"{o}m, Olsson, Osterblom, Scheffer,
  and Worm}]{Berkes2006}
Berkes, F., Hughes, T.~P., Steneck, R.~S., Wilson, J.~A., Bellwood, D.~R.,
  Crona, B., Folke, C., Gunderson, L.~H., Leslie, H.~M., Norberg, J.,
  Nystr\"{o}m, M., Olsson, P., Osterblom, H., Scheffer, M., Worm, B., Mar.
  2006. {Ecology. Globalization, roving bandits, and marine resources.} Science
  (New York, N.Y.) 311~(5767), 1557--8.
\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/16543444}

\bibitem[{Boettiger and Hastings(2012)}]{Boettiger2012b}
Boettiger, C., Hastings, A., 2012. {Quantifying Limits to Detection of Early
  Warning for Critical Transitions}. Journal Of The Royal Society Interface.

\bibitem[{{California Supreme Court}(1968)}]{PeopleCollins1968}
{California Supreme Court}, 1968. People v. collins.

\bibitem[{Carpenter(2011)}]{Carpenter2011}
Carpenter, J., Feb. 2011. {May the Best Analyst Win}. Science 331~(6018),
  698--699.
\newline\urlprefix\url{http://www.sciencemag.org/cgi/doi/10.1126/science.331.6018.698}

\bibitem[{Carpenter and Brock(2006)}]{Carpenter2006}
Carpenter, S.~R., Brock, W.~A., 2006. {Rising variance: a leading indicator of
  ecological transition.} Ecology letters 9~(3), 311--8.
\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/16958897}

\bibitem[{Courchamp et~al.(2008)Courchamp, Berec, and
  Gascoigne}]{Courchamp2008}
Courchamp, F., Berec, L., Gascoigne, J., 2008. {Allee Effects in Ecology and
  Conservation}. Oxford University Press, USA.
\newline\urlprefix\url{http://www.amazon.com/Effects-Ecology-Conservation-Franck-Courchamp/dp/0198570309}

\bibitem[{Dai et~al.(2012)Dai, Vorselen, Korolev, and Gore}]{Dai2012}
Dai, L., Vorselen, D., Korolev, K.~S., Gore, J., May 2012. {Generic Indicators
  for Loss of Resilience Before a Tipping Point Leading to Population
  Collapse}. Science 336~(6085), 1175--1177.
\newline\urlprefix\url{http://www.sciencemag.org/cgi/doi/10.1126/science.1219805}

\bibitem[{Dakos et~al.(2011)Dakos, K\'{e}fi, Rietkerk, Nes, and
  Scheffer}]{Dakos2011}
Dakos, V., K\'{e}fi, S., Rietkerk, M., Nes, E. H.~V., Scheffer, M., 2011.
  {Slowing Down in Spatially Patterned Ecosystems at the Brink of Collapse}.
  The American Naturalist.

\bibitem[{Dakos et~al.(2008)Dakos, Scheffer, van Nes, Brovkin, Petoukhov, and
  Held}]{Dakos2008}
Dakos, V., Scheffer, M., van Nes, E.~H., Brovkin, V., Petoukhov, V., Held, H.,
  Sep. 2008. {Slowing down as an early warning signal for abrupt climate
  change.} Proceedings of the National Academy of Sciences 105~(38), 14308--12.
\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/18787119}

\bibitem[{Ditlevsen and Johnsen(2010)}]{Ditlevsen2010}
Ditlevsen, P.~D., Johnsen, S.~J., Oct. 2010. {Tipping points: Early warning and
  wishful thinking}. Geophysical Research Letters 37~(19), 2--5.
\newline\urlprefix\url{http://www.agu.org/pubs/crossref/2010/2010GL044486.shtml}

\bibitem[{Drake and Griffen(2010)}]{Drake2010}
Drake, J.~M., Griffen, B.~D., Sep. 2010. {Early warning signals of extinction
  in deteriorating environments}. Nature 467~(7314), 456--459.
\newline\urlprefix\url{http://www.nature.com/doifinder/10.1038/nature09389}

\bibitem[{Guttal and Jayaprakash(2008)}]{Guttal2008}
Guttal, V., Jayaprakash, C., Dec. 2008. {Spatial variance and spatial skewness:
  leading indicators of regime shifts in spatial ecological systems}.
  Theoretical Ecology 2~(1), 3--12.
\newline\urlprefix\url{http://www.springerlink.com/index/10.1007/s12080-008-0033-1}

\bibitem[{Hirota et~al.(2011)Hirota, Holmgren, {Van Nes}, and
  Scheffer}]{Hirota2011}
Hirota, M., Holmgren, M., {Van Nes}, E.~H., Scheffer, M., Oct. 2011. {Global
  Resilience of Tropical Forest and Savanna to Critical Transitions}. Science
  334~(6053), 232--235.
\newline\urlprefix\url{http://www.sciencemag.org/cgi/doi/10.1126/science.1210657}

\bibitem[{K\'{e}fi et~al.(2007)K\'{e}fi, Rietkerk, Alados, Pueyo, Papanastasis,
  Elaich, and de~Ruiter}]{Kefi2007}
K\'{e}fi, S., Rietkerk, M., Alados, C.~L., Pueyo, Y., Papanastasis, V.~P.,
  Elaich, A., de~Ruiter, P.~C., Sep. 2007. {Spatial vegetation patterns and
  imminent desertification in Mediterranean arid ecosystems.} Nature
  449~(7159), 213--7.
\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/17851524}

\bibitem[{Lenton(2011)}]{Lenton2011}
Lenton, T.~M., Jun. 2011. {Early warning of climate tipping points}. Nature
  Climate Change 1~(4), 201--209.
\newline\urlprefix\url{http://www.nature.com/doifinder/10.1038/nclimate1143}

\bibitem[{Lenton et~al.(2012)Lenton, Livina, and Dakos}]{Lenton2012}
Lenton, T.~M., Livina, V., Dakos, V., 2012. {Early warning of climate tipping
  points from critical slowing down: comparing methods to improve robustness}.
  Philosophical Transactions of The Royal Society A (in press).
\newline\urlprefix\url{http://sites.google.com/site/noraclpreprints/contributions/lenton-et-al/Early\_warning\_methods\_revised\_with\_figs\_and\_appendix2.pdf}

\bibitem[{Livina and Lenton(2007)}]{Livina2007}
Livina, V.~N., Lenton, T.~M., Feb. 2007. {A modified method for detecting
  incipient bifurcations in a dynamical system}. Geophysical Research Letters
  34~(3), 1--5.
\newline\urlprefix\url{http://www.agu.org/pubs/crossref/2007/2006GL028672.shtml}

\bibitem[{Mumby et~al.(2007)Mumby, Hastings, and Edwards}]{Mumby2007}
Mumby, P.~J., Hastings, A., Edwards, H.~J., Nov. 2007. {Thresholds and the
  resilience of Caribbean coral reefs.} Nature 450~(7166), 98--101.
\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/17972885}

\bibitem[{Scheffer et~al.(2009)Scheffer, Bascompte, Brock, Brovkin, Carpenter,
  Dakos, Held, van Nes, Rietkerk, and Sugihara}]{Scheffer2009}
Scheffer, M., Bascompte, J., Brock, W.~A., Brovkin, V., Carpenter, S.~R.,
  Dakos, V., Held, H., van Nes, E.~H., Rietkerk, M., Sugihara, G., 2009.
  {Early-warning signals for critical transitions.} Nature 461~(7260), 53--9.
\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/19727193}

\bibitem[{Scheffer et~al.(2001)Scheffer, Carpenter, Foley, Folke, and
  Walker}]{Scheffer2001}
Scheffer, M., Carpenter, S.~R., Foley, J.~A., Folke, C., Walker, B., Oct. 2001.
  {Catastrophic shifts in ecosystems.} Nature 413~(6856), 591--6.
\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/11595939}

\bibitem[{Thompson and Sieber(2010)}]{Thompson2010}
Thompson, J. M.~T., Sieber, J., Dec. 2010. {Climate tipping as a noisy
  bifurcation: a predictive technique}. IMA Journal of Applied Mathematics
  76~(1), 27--46.
\newline\urlprefix\url{http://imamat.oxfordjournals.org/cgi/doi/10.1093/imamat/hxq060}

\bibitem[{Thompson and Schumann(1987)}]{Thompson1987}
Thompson, W.~C., Schumann, E.~L., 1987. {Interpretation of statistical evidence
  in criminal trials: The prosecutor's fallacy and the defense attorney's
  fallacy.} Law and Human Behavior 11~(3), 167--187.
\newline\urlprefix\url{http://psycnet.apa.org/journals/lhb/11/3/167/
  http://doi.apa.org/getdoi.cfm?doi=10.1007/BF01044641}

\bibitem[{van Nes and Scheffer(2007)}]{vanNes2007}
van Nes, E.~H., Scheffer, M., Jun. 2007. {Slow recovery from perturbations as a
  generic indicator of a nearby catastrophic shift.} The American naturalist
  169~(6), 738--47.
\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/17479460}

\bibitem[{Veraart et~al.(2011)Veraart, Faassen, Dakos, van Nes, L\"{u}rling,
  and Scheffer}]{Veraart2011}
Veraart, A.~J., Faassen, E.~J., Dakos, V., van Nes, E.~H., L\"{u}rling, M.,
  Scheffer, M., Dec. 2011. {Recovery rates reflect distance to a tipping point
  in a living system}. Nature, 2--5.
\newline\urlprefix\url{http://www.nature.com/doifinder/10.1038/nature10723}

\end{thebibliography}

 \end{document}
